Climate Sensitivity of Tropical Tree Growth

### 2024-06-25 scripts/explore_analysis.Rmd
data summary plots

## Climate Sensitivity of Tropical Tree Growth

@@

#reading the GitHub PAT (stored privately and locally)
pat<-as.character(read.table("git_pat.txt")[1,1])

#remotes:::download("DendroByBand.RData",
#                   "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
#                   auth_token = pat)

load("DendroByBand.RData")

@@

req <- GET("https://api.github.com/repos/forestgeo/Climate/git/trees/master?recursive=1")
stop_for_status(req)
filelist <- unlist(lapply(content(req)$tree, "[", "path"), use.names = F)
path<-grep("Climate_Data/CRU/CRU_v4_06", filelist, value = TRUE, fixed = TRUE)
datacsvs<-grep("pre|tmx", path, value = T)

#datacsvs<-datacsvs[!grepl("BLANK", datacsvs)]
datacsvs<-paste0("https://github.com/forestgeo/Climate/raw/master/", datacsvs)

start_date<-min(hkk_OBSID$year,na.rm=T)
end_date<-max(hkk_OBSID$year,na.rm=T)

climate_data_orig<-bind_rows(datacsvs %>%
                               read_csv(na = c("", "NA", "N/A", "NULL"), show_col_types = FALSE))%>%
  clean_names() %>%
  filter(sites_sitename %in% c("HKK",
                               "Huai Kha Khaeng")) %>%
  select(-c(sites_sitename, n_obs)) %>%
  rename(year=yr,
         month=mo)%>%
  filter(start_date-1 <= year & year <= end_date) # limiting to specified date range but extending to a year before

@@

pre_yr_census<-ggplot()+
  geom_area(data=climate_data_orig[climate_data_orig$variable=="pre",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value), fill="blue", alpha=0.5)+
  geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
  scale_y_continuous(    # Features of the first axis
    name = "monthly rainfall (mm)",

    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("Date")+theme_bw()+
  ggtitle("HKK dendroband censuses and CRU climate data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

@@

@@

@@

@@

@@

@@

>>>>

@@

#reading the GitHub PAT (stored privately and locally)
pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])

# remotes:::download("DendroByBand.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
#                    auth_token = pat)

load("../data/HKK-dendro/DendroByBand.RData")

@@

#CHIRPS data ----

# CHIRPS data GEE code here - https://code.earthengine.google.com/894597bd093906fb0ff82320f29c0842

chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
colnames(chirps_hkk)[1]<-"time"

#making dates manageable
chirps_hkk<- chirps_hkk %>%
  dplyr::mutate(time=as.Date(time, format = "%b %d, %Y"),
                year=year(time),
                month=month(time))

@@

pre_yr_census<-ggplot()+
  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
  # scale_y_continuous(    # Features of the first axis
  #   name = "monthly rainfall (mm)",
  #   
  #   # Add a second axis and specify its features
  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("Date")+theme_bw()+
  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

@@

#rainfall per year

#-----------------Decisions on climate data-----------------------=
# Summarising climate data to relevant time window for SCBI
# As per issue #42 and KAT pub https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.15934
# HKK best climate variables are:
# PPT - previous Sep to current June (it is actually precipitation day freq, but using sum now)
# Tmax - current Apr to Oct
#---------------------------------------------------------------=

#applying these decisions and rescaling

#mean annual precip from 1981 to 2023 is 1698.4

#code - https://code.earthengine.google.com/b0eda787f2e1ba578bbf50a8a9d04843

climate_data_yr<-chirps_hkk %>%
  group_by(year) %>%
  summarise(
    tot_precip=sum(precipitation),
    pre_SeptoDec=sum(precipitation[month>=9]), #Sep to Dec precip for each year
    pre_JantoJun=sum(precipitation[month<=6]) #Jan to June precip for each year
    #tmx_window=sum(value[variable == "tmx"& month %in% 4:10])
)%>% #sum of potential evapotranspiration for current May to July
  mutate(
    pre_window=pre_JantoJun+lag(pre_SeptoDec), #adding prev year's June to Dec and current year's Jan to June
    tot_pre_diff=tot_precip-1698.4
    )

#standard deviation is 254.06

precip_plot<-ggplot(climate_data_yr)+
  geom_bar(aes(x=year, y=tot_pre_diff), stat="identity")+
  ylab("difference from 40-year mean (mm)")+
  geom_hline(yintercept = c(254.06, -254.06), lty=2)+
  theme_bw()
precip_plot

# library
library(ggridges)
library(ggplot2)
library(viridis)
library(hrbrthemes)

chirps_hkk$year_fac<-as.character(chirps_hkk$year)
chirps_hkk$year_fac<-factor(chirps_hkk$year)

# Plot
ggplot(chirps_hkk, aes(x = precipitation, y = year_fac, 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "rainfall", option = "C") +
  labs(title = 'Rainfall in HKK') +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

@@

# species_all<-dendro.data.noouts %>%
#   group_by(sp, treeID) %>%
#   dplyr::summarise(n.obs=length(treeID))%>%
#   ungroup() %>%
#   group_by(sp) %>%
#   dplyr::summarise(n.tree=length(treeID),
#                    n.obs.tot=sum(n.obs))%>%
#   ungroup()%>%
#   arrange(-n.tree)
# 
# write.csv(species_all, "../data/HKK-dendro/species.csv", row.names = F)

@@

```{r echo=FALSE, eval=F}

@@

```{r maps of species with tree ring records, echo=F, eval=F}
library(ggplot2)

@@

## Detrending all species

@@

```
## Fastest and slowest in 2015 and 2016

dendro.data.noouts <- dendro.data.noouts %>%
  group_by(year) %>% 
  dplyr::mutate(growth_quantile = ntile(diainc, 4),
                resid_quantile = ntile (inc.resid, 4))

growth_2015 <- dendro.data.noouts %>%
  filter(year==2015)

quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()

growth_2014 <- dendro.data.noouts %>%
  filter(year==2014)

quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2016 <- dendro.data.noouts %>%
  filter(year==2016)

quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()

growth_2017 <- dendro.data.noouts %>%
  filter(year==2017)

quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))

growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
  geom_density(fill="brown", col="brown", alpha=0.5) +
  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()

growth_2014_2015_plot<-ggplot() +
  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
  geom_vline(xintercept=quantiles_2015, col="brown")+ 
  xlab("diameter increment (cm)") + ylab("density")+
  theme_bw()

dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))

growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
  labs(title = 'tree growth in HKK (cm)', y="year") +
  #theme_ipsum() +
  geom_vline(xintercept=0, lty=2)+
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

growth_all_years

png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
growth_2014_2015_plot
dev.off()

png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
growth_all_years
dev.off()

##Distribution of data with size, exposure and deciduousness

#2014 data distribution with size, exposure and deciduousness

png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

length(unique(growth_2014$sp))

png("../doc/display/growth_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = diainc, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'growth in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
dev.off()
